---
title: "06_HIV_Behavior"
output:
  word_document: default
  html_document: default
---

# HIV Behavior - National South Africa

## Load Libraries

```{r}
# Data manipulation
library(dplyr)
library(tidyr)
library(readr)
library(here)
library(purrr)

# Visualization and summaries
library(ggplot2)
library(skimr)
library(visdat)
```

## Load Data
```{r}
# Load the HIV behavior dataset
hiv_df <- read_csv(
  here("data", "raw", "hiv-behavior_national_zaf.csv"),
  col_names = TRUE,   # use first row as column names
  col_types = cols()  # suppress guessing messages
)

# Step 2: Remove first row if it contains metadata
hiv_df <- hiv_df[-1, ]

# Step 3: Reset row names
rownames(hiv_df) <- NULL

cat("HIV behavior dataset loaded successfully.\n")
cat("Dimensions:", dim(hiv_df), "\n")
```

## Initial Assessment 
```{r}
# Quick glimpse
glimpse(hiv_df)

# Summary of missingness
skim(hiv_df)

# Visualize missing values
vis_miss(hiv_df)


```


## Handle Duplicates

```{r}
# Check for exact duplicates
cat("Exact duplicates:", sum(duplicated(hiv_df)), "\n")

# Remove exact duplicates
hiv_df <- hiv_df %>% distinct()
cat("Dimensions after duplicate removal:", dim(hiv_df), "\n")


```
## Covert Data Types
```{r}
# Convert numeric columns safely
num_cols <- c("value", "precision", "denominator_weighted", "denominator_unweighted",
              "ci_low", "ci_high", "survey_year", "indicator_order",
              "characteristic_id", "characteristic_order", "survey_year_label")
num_cols <- num_cols[num_cols %in% colnames(hiv_df)] # only existing columns

hiv_df <- hiv_df %>%
  mutate(across(all_of(num_cols), as.numeric))

# Logical columns
logical_cols <- c("is_total", "is_preferred")
logical_cols <- logical_cols[logical_cols %in% colnames(hiv_df)]
hiv_df <- hiv_df %>% mutate(across(all_of(logical_cols), ~as.logical(as.integer(.))))

# Check structure
str(hiv_df)

```
- Numeric columns (value, precision, denominator_weighted, denominator_unweighted, ci_low, ci_high, survey_year, indicator_order, characteristic_id, characteristic_order, survey_year_label) were converted to numeric.

- Logical columns (is_total, is_preferred) were converted to boolean values.

- Conversion ensures accurate calculations and proper visualization.
## Handle Missing Values

```{r}
# Impute survey_year_label with survey_year if missing
if ("survey_year_label" %in% colnames(hiv_df)) {
  hiv_df <- hiv_df %>%
    mutate(survey_year_label = ifelse(is.na(survey_year_label), survey_year, survey_year_label))
}

# Impute survey_type with "Unknown" if missing
if ("survey_type" %in% colnames(hiv_df)) {
  hiv_df <- hiv_df %>%
    mutate(survey_type = ifelse(is.na(survey_type), "Unknown", survey_type))
}

# Recalculate missing summary
missing_summary <- data.frame(
  Column = colnames(hiv_df),
  n_missing = colSums(is.na(hiv_df)),
  total_rows = nrow(hiv_df),
  missing_percent = round(colSums(is.na(hiv_df))/nrow(hiv_df)*100, 2)
)

# Impute denominators with median of available values
hiv_df <- hiv_df %>%
  mutate(
    DenominatorWeighted = ifelse(is.na(DenominatorWeighted),
                                 median(DenominatorWeighted, na.rm = TRUE),
                                 DenominatorWeighted),
    DenominatorUnweighted = ifelse(is.na(DenominatorUnweighted),
                                   median(DenominatorUnweighted, na.rm = TRUE),
                                   DenominatorUnweighted)
  )

# Function to calculate mode
get_mode <- function(x) {
  ux <- unique(x[!is.na(x)])
  ux[which.max(tabulate(match(x, ux)))]
}

# Impute missing values with most frequent value
hiv_df <- hiv_df %>%
  mutate(
    DHS_CountryCode = ifelse(is.na(DHS_CountryCode), get_mode(DHS_CountryCode), DHS_CountryCode),
    IndicatorOrder = ifelse(is.na(IndicatorOrder), get_mode(IndicatorOrder), IndicatorOrder),
    IndicatorType = ifelse(is.na(IndicatorType), get_mode(IndicatorType), IndicatorType),
    CharacteristicId = ifelse(is.na(CharacteristicId), get_mode(CharacteristicId), CharacteristicId),
    CharacteristicOrder = ifelse(is.na(CharacteristicOrder), get_mode(CharacteristicOrder), CharacteristicOrder),
    CharacteristicCategory = ifelse(is.na(CharacteristicCategory), get_mode(CharacteristicCategory), CharacteristicCategory),
    CharacteristicLabel = ifelse(is.na(CharacteristicLabel), get_mode(CharacteristicLabel), CharacteristicLabel),
    IsTotal = ifelse(is.na(IsTotal), get_mode(IsTotal), IsTotal),
    IsPreferred = ifelse(is.na(IsPreferred), get_mode(IsPreferred), IsPreferred),
    SDRID = ifelse(is.na(SDRID), get_mode(SDRID), SDRID),
    SurveyYearLabel = ifelse(is.na(SurveyYearLabel), get_mode(SurveyYearLabel), SurveyYearLabel),
    SurveyType = ifelse(is.na(SurveyType), get_mode(SurveyType), SurveyType)
  )
# Drop columns that are 100% missing
cols_to_drop <- c("ByVariableLabel", "RegionId", "CILow", "CIHigh", "LevelRank")
cols_to_drop <- intersect(cols_to_drop, colnames(hiv_df))  # only if they exist

hiv_df <- hiv_df %>% select(-all_of(cols_to_drop))
cat("Dropped completely missing columns:\n")
print(cols_to_drop)

# Verify that missing values are handled
colSums(is.na(hiv_df))


```
Handling Missing Values

Strategies applied:

1. Survey Year Label: Filled missing survey_year_label with survey_year.

2. Survey Type: Filled missing survey_type with "Unknown".

3. Denominator columns: Filled with median of available values.

4. Categorical columns: Filled missing values with the mode (most frequent value).

5. Dropped columns that were 100% missing (ByVariableLabel, RegionId, CILow, CIHigh, LevelRank).

6. Outcome: No missing values remain, ensuring the dataset is analysis-ready.
## Handle Outliers

```{r}
# Winsorize HIV Behavior 'Value' at 1st and 99th percentiles

# First, check the structure and type of Value column
cat("Structure of Value column:\n")
str(hiv_df$Value)

cat("\nClass of Value column:", class(hiv_df$Value), "\n")
cat("First few values:", head(hiv_df$Value), "\n")

# Check for any non-numeric values
cat("\nNon-numeric values in Value column:\n")
print(hiv_df$Value[!is.na(hiv_df$Value) & !is.numeric(hiv_df$Value)])

# Convert to numeric if necessary (handling any character values)
hiv_df$Value <- as.numeric(as.character(hiv_df$Value))

# Check for NAs introduced by conversion
cat("\nNA values after conversion:", sum(is.na(hiv_df$Value)), "\n")

# Now proceed with winsorization
lower_val <- quantile(hiv_df$Value, 0.01, na.rm = TRUE)
upper_val <- quantile(hiv_df$Value, 0.99, na.rm = TRUE)

cat("\n1st percentile (lower bound):", lower_val, "\n")
cat("99th percentile (upper bound):", upper_val, "\n")

hiv_df <- hiv_df %>%
  mutate(
    Value = pmax(pmin(Value, upper_val), lower_val)
  )

# Create log transformation (using log1p to handle zeros)
hiv_df <- hiv_df %>%
  mutate(Value_log = log1p(Value))

# Check summary
cat("\nSummary of Value after winsorization:\n")
summary(hiv_df$Value)

cat("\nSummary of log-transformed Value:\n")
summary(hiv_df$Value_log)

# Visualize with boxplot
ggplot(hiv_df, aes(y = Value)) +
  geom_boxplot(fill = "steelblue", outlier.color = "red", alpha = 0.7) +
  labs(
    title = "Boxplot of HIV Behavior 'Value' After Winsorization",
    y = "Value",
    caption = "Winsorized at 1st and 99th percentiles"
  ) +
  theme_minimal()

# Additional visualization for log-transformed values
ggplot(hiv_df, aes(y = Value_log)) +
  geom_boxplot(fill = "lightgreen", outlier.color = "red", alpha = 0.7) +
  labs(
    title = "Boxplot of Log-Transformed HIV Behavior 'Value'",
    y = "log(Value + 1)",
    caption = "log1p transformation applied"
  ) +
  theme_minimal()


```
Handling Outliers

- The numeric column Value was Winsorized at the 1st and 99th percentiles to reduce the influence of extreme values.

- Log transformation (log1p) was applied to Value to normalize distributions and handle zero values.

- Boxplots were created to visualize both the Winsorized and log-transformed values.


```{r}


# Quick check of structure and summary
str(hiv_df)
summary(hiv_df)

# Check for any remaining NAs
colSums(is.na(hiv_df))

# Save cleaned dataset
write_csv(hiv_df, here("data", "processed", "hiv-behavior_national_zaf_clean.csv"))

cat("HIV behavior dataset cleaned and saved to data/processed/hiv-behavior_national_zaf_clean.csv\n")

```





